Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-01-07 sources/bmc_template_submission/bmc_article.tex
intro small edits

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with with molecular sequence data that has never been 
incorporated into any phylogenetic estimate provides the opportunity to capture the evolutionary history of many taxa.
Here, we introduce Physcraper, a tool that automates database connections
to build on published alignments to extend phylogenetic inferences with more taxa.
Physcraper can be used to to update a starting tree and single locus alignments
with public DNA data, and link the tips in these trees to a unified, interoperable taxonomic resource \cite{rees2017automated}.

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases which constitute an amazing
resource for biological discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of biological data across databases are varying taxonomic idiosyncrasies. 
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020}, and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019} among others.
By using the existing OpenTree taxonomy programmatic tools to map tip names,
Physcraper has a framework for connecting updated phylogenies data from any
biological database.

@@

Assembling a DNA alignment from such a massive database can be done "by hand",
but that is a time consuming approach which is not highly reproducible.
A variety of computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been applied to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

@@

Physcraper is a Python pipeline using OpenTree's programmatic access
protocols (API's) to automatically
links phylogenies mapped to a standardized taxonomy \cite{mctavish2015phylesystem}, to alignments
from TreeBASE \cite{sanderson1994treebase}, data from GenBank \cite{benson2000genbank}.
Physcraper provides automated and reproducible workflows for updating existing phylogenetic estimates.

>>>>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with molecular sequence data that has never been
incorporated into any phylogenetic estimate provides the opportunity to capture
the evolutionary history of many taxa.
Here, we introduce Physcraper, a tool that automates database connections
to build on published alignments to extend phylogenetic inferences with more taxa.
Physcraper can be used to update a starting tree and single locus alignments
with public DNA data, and link the tips in these trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

Data such as
geographical location, fossil ranges, and genetic and phenotypic information increasingly
available in public databases constitute an amazing
resource for biological discovery \cite{baxevanis2015importance}.
One of the main challenges for automatic
integration of biological data across databases are varying taxonomic idiosyncrasies.
To address this challenge, the Open Tree of Life project (OpenTree)
created a unified taxonomy for name standardization, by integrating taxonomic
data from several databases \cite{rees2017automated}, including the USA National
Center for Biodiversity Information (NCBI) taxonomy \cite{federhen_ncbi_2012, schoch_ncbi_2020},
and the Global Biodiversity Information Facility (GBIF) \cite{gbif_secretariat_gbif_2019} among others.
By using the existing OpenTree taxonomy programmatic tools to map tip names,
Physcraper has a framework for connecting updated phylogenies data from any
biological database.

@@

Assembling a DNA alignment from such a massive database can be done "by hand",
but that is a time consuming approach which is not highly reproducible.
A variety of computational pipelines that mine DNA databases fast, efficiently, and reproducibly
have been developed and used to infer phylogenetic relationships in a variety of organisms
(e.g., \cite{smith2009mega, antonelli2017toward, izquierdo2014pumper}).
While genomics has, and will continue to, revolutionize phylogenetic inference,
the diversity of alternative genomic sequencing approaches implemented produce
largely non-overlapping homology hypotheses across taxa \cite{jones2016targeted},
creating challenges for phylogenetic reconstruction.
Phylogenomics addresses this problem by focusing on targeted capture of informative
regions \cite{andermann2020guide}.
However, fine-grained curated markers and alignments can significantly improve phylogenetic
reconstructions, even in phylogenomic analyses \cite{fragoso2017pilot}.

@@

Physcraper is a Python pipeline using OpenTree's programmatic access
protocols (API's) to automatically
link phylogenies mapped to a standardized taxonomy \cite{mctavish2015phylesystem}, to alignments
from TreeBASE \cite{sanderson1994treebase}, data from GenBank \cite{benson2000genbank}.
Physcraper provides automated and reproducible workflows for updating existing phylogenetic estimates.
